---
title: "Peer Graded Assignment - Practical Machine Learning"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

Using data from use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants, this analysis is aimed to predict the "classe" variable from the 20 test cases from the test dataset.

The training data for this project are available here:

<https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv>

The test data are available here:

<https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv>

Dataset: [Human Activity Recognition Dataset](http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har)

## Methods

The analysis were done by cleaning irrelevant variable (X, username, timestamps, and window) from both the training and the test data. The cleaned training data were then used to training and modelling algorithms. The training data were splitted into the data used for training the model, and validation, which were used to build Gradient Boosting Machine(GBM) and Random Forest (RF). Accuracy, Kappa Statistics, and Out-of-sample errors were attained from Confusion Matrix. The best model were then used to predict the testing dataset

## Load the data

```{r}
dir <- "C:\\Users\\Regina Maria\\OneDrive\\Desktop"
setwd(dir)
getwd()
```

```{r}
fileurl_train <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
fileurl_test <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

#download the file and verifying the file
if (!file.exists("data")){
    dir.create("data")
}

download.file(fileurl_train, destfile = "./data/train.csv", method = "curl")
download.file(fileurl_test, destfile = "./data/test.csv", method = "curl")

list.files("./data")

dateDownloaded <- date()
dateDownloaded
```

## Reading and Cleaning the data

```{r}
#read the train and test data
traindata <- read.csv("./data/train.csv", sep = ",", header = TRUE)
testdata <- read.csv("./data/test.csv", sep = ",", header = TRUE)

#removing NAs
traindata <- traindata[, colSums(is.na(traindata)) == 0]
testdata <- testdata[, colSums(is.na(testdata)) == 0]

#removing near zero variance
library(caret)
nzv_train <- nearZeroVar(traindata)
nzv_test <- nearZeroVar(testdata)
traindata <- traindata[, -nzv_train]
testdata <- testdata[, -nzv_test]

#removing irrelevant data for model training
variables <- c("X", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp", "num_window")
traindata <- traindata[, !(names(traindata) %in% variables)]
testdata <- testdata[, !(names(testdata) %in% variables)]

#making classe into factor
traindata$classe <- as.factor(traindata$classe)

head(traindata)
head(testdata)
```

## Separating training data for training and validation

```{r}
#creating data partition
set.seed(3233)
inTrain <- createDataPartition(y = traindata$classe, p = 0.75, list = FALSE)
training <- traindata[inTrain,]
validation <- traindata[-inTrain,]
dim(training); dim(validation)
```

```{r}
#creating kfolds
folds <- createFolds(y = training$classe, k = 10, list = TRUE, returnTrain = TRUE)
sapply(folds, length)

folds <- createFolds(y = validation$classe, k = 10, list = TRUE, returnTrain = TRUE)
sapply(folds, length)
```

## Fitting GBM model

```{r}
#gbm - boosting
set.seed(3233)
controlgbm <- trainControl(method="cv", 5, allowParallel = TRUE)
gbmmodel <- train(classe~., data = training, method = "gbm", verbose = FALSE, trControl = controlgbm)
gbmmodel
```

```{r}
#confusion matrix for validating model
predictgbm <- predict(gbmmodel, validation)
confusionMatrix(validation$classe, predictgbm)
```

## Fitting RF model

```{r}
#rf 
set.seed(3233)
controlrf <- trainControl(method="cv", 5, allowParallel = TRUE)
rfmodel <- train(classe~., data = training, method = "rf", trControl = controlrf)
rfmodel
```

```{r}
#confusion matrix for validating model
predictrf <- predict(rfmodel, validation)
confusionMatrix(validation$classe, predictrf)
```

```{r}
#out of sample error random forest
matrixgbm <- confusionMatrix(validation$classe, predictgbm)
accuracygbm <- matrixgbm$overall["Accuracy"]
errorgbm <- 1 - as.numeric(accuracygbm)
errorgbm
```

```{r}
#out of sample error random forest
matrixrf <- confusionMatrix(validation$classe, predictrf)
accuracyrf <- matrixrf$overall["Accuracy"]
errorrf <- 1 - as.numeric(accuracyrf)
errorrf
```

## Comparing model

```{r}
#creating data frame
model_compare <- data.frame(
    Model = c("GBM", "RF"),
    Accuracy = c(as.numeric(accuracygbm), as.numeric(accuracyrf)),
    OutOfSampleError = c(as.numeric(errorgbm), as.numeric(errorrf))
)

#creating bar plot for comparison
par(mfrow = c(1,2))

barplot(model_compare$Accuracy,
        names.arg = model_compare$Model,
        main = "Model Accuracy",
        ylab = "Accuracy",
        col = "blue")

barplot(model_compare$OutOfSampleError,
        names.arg = model_compare$Model,
        main = "Out-of-Sample Error",
        ylab = "Error",
        col = "red")
```

Higher kappa statistic and accuracy showed Random Forest model showed to be a better model than boosted model. Random forest also showed to have lower out of sample error. Hence, **Random Forest were used as the final model for prediction**

## Predicting the testing model

```{r}
finalmodel <- rfmodel
predicttest <- predict(finalmodel, testdata)
table(predicttest)
```
